Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-10-23 _tex/index.tex
Built site for gh-pages

Nevertheless, despite the growing number of CJ studies, the unsystematic
and fragmented research approaches employed in the literature have
overlooked several critical issues concerning the method. These issues
fall into three main categories: the disconnect between CJ's structural
and measurement model, the over-reliance on Thurstone's Case 5
\citeyearpar{Thurstone_1927} in its measurement model, and the role of
the comparison algorithms and the number of comparisons on the method's
reliability and validity. In the following sections, each issue will be
discussed in detail, followed by the introduction of a theoretical model
and its translation into a statistical model that addresses all three
concerns simultaneously.

\section{Overlooked issues in CJ literature}\label{sec-theory-issues}

\subsection{The disconnect between structural and measurement
model}\label{the-disconnect-between-structural-and-measurement-model}

In the CJ literature, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model
for CJ because it specifies how latent variables are estimated from
observed ones \citep{Hoyle_et_al_2023, Kline_et_al_2023}. In a CJ study,
multiple judges conduct several rounds of pairwise comparisons to
evaluate the relative manifestation of a trait between a pair of
stimuli. This evaluation results in a dichotomous outcome that indicates
which stimulus is perceived to exhibit a higher degree of the trait. The
BTL model then uses these observed outcomes to estimate scores that
represent the latent trait of interest
\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.

Moreover, in the CJ literature, it is common practice for BTL-generated
scores, or their transformations, to undergo additional data analysis
and hypothesis testing. Researchers have utilized these scores in
independent analyses to identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

\subsection{Case 5 and the measurement
model}\label{case-5-and-the-measurement-model}

\subsection{The efficiency and reliability of comparison
algorithms}\label{the-efficiency-and-reliability-of-comparison-algorithms}

\subsection{What is the solution}\label{what-is-the-solution}

@@

\subsection{A theoretical model for the CJ}\label{sec-theory-scientific}

\subsection{From theory to statistical
model}\label{sec-theory-statmodel}

>>>>

Nevertheless, despite the growing number of studies on CJ, unsystematic
and fragmented research approaches in the literature have left several
critical issues unaddressed. These issues can be grouped into three main
categories: the disconnect between CJ's structural and measurement
model, the over-reliance on the assumptions of Thurstone's Case 5
\citeyearpar{Thurstone_1927} in its measurement model, and the unclear
role and impact of comparison algorithms and the number of comparisons
on the method's reliability and validity. The following sections will
discuss each of these issues in detail,, followed by the introduction of
a theoretical model and its statistical translation, which aim to
address all three concerns simultaneously.

\section{The three critical issues in CJ
literature}\label{sec-theory-issues}

\subsection{The disconnect between structural and measurement
models}\label{the-disconnect-between-structural-and-measurement-models}

In a typical CJ study, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} serves as the measurement model.
This model specifies how latent variables are estimated from observed
ones \citep{Hoyle_et_al_2023, Kline_et_al_2023}. In the context of CJ,
multiple judges engage in several rounds of pairwise comparisons to
assess the relative manifestation of a trait between two stimuli. Each
comparison generates a dichotomous outcome, indicating which stimulus is
perceived to exhibit a higher degree of the trait. The BTL model then
uses these observed outcomes to estimate scores that represent the
latent trait of interest
\citep{Pollitt_2012a, Pollitt_2012b, Whitehouse_2012, Jones_2015, vanDaal_et_al_2016, Lesterhuis_2018, Boonen_et_al_2020, Bouwer_et_al_2023}.

Moreover, researchers often use these BTL-generated scores or their
transformations separately, applying additional analyses or hypothesis
testing. The CJ literature shows that these scores have been employed to
identify `misfit' judges and stimuli
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018}, detect
biases in judges' ratings \citep{Pollitt_et_al_2003, Pollitt_2012b},
calculate correlations with other scoring methods
\citep{Goossens_et_al_2018, Bouwer_et_al_2023}, or test hypotheses
related to the trait of interest
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.

However, the statistical literature warns against using estimated scores
for separate analyses and hypothesis testing. BTL-generated scores are
parameter estimates that inherently carry uncertainty. Neglecting this
uncertainty can inflate the precision and statistical power in
independent analyses and tests. This oversight, in turn, may increase
the risk of committing a type I error, where the null hypothesis is
incorrectly rejected \citep{McElreath_2020}. To mitigate this risk,
principles from Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023, Kline_et_al_2023} recommend conducting such
analyses within the structural model of CJ rather than separately.
Therefore, an integrated approach combining both structural and
measurement models is advisable.

\subsection{The assumptions of Case 5 and the measurement
model}\label{the-assumptions-of-case-5-and-the-measurement-model}

\subsection{The role and impact of comparison
algorithms}\label{the-role-and-impact-of-comparison-algorithms}

@@

\subsection{A theoretical model for CJ}\label{sec-theory-theoretical}

\subsection{From theory to statistics}\label{sec-theory-statistics}
